Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-05 main_new.tex
updates

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem phenomena (e.g. blooms, anoxic
zones, plumes) all with significant impact both directly and
indirectly on society. Reverberating the requirement of more effective
ocean sensing practices, Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observational efforts sometimes leading to severe environmental
problems such as the recent (May 2018) outbreak of harmful algal
blooms in Northern Norway\footnote{Report from the Institute of Marine
  Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

@@

Our attention in this work is towards spatial characterization of a
frontal system, and more specifically fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) taken as cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind and temperature differences, this boundary in water masses often
gets distorted. Prior knowledge about the location and evolution of
these features is therefore highly uncertain, making deterministic
planning challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $10 m^2$ -- $5 km^2$ where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit scope to cover only lateral spatial
elements ([north, east] domain). Moreover, for this case, time-effects
can be regarded as static when acquiring AUV data over a short time
window and limited region. However, the methodological framework can
be extended to higher dimensional situations in a similar manner.

@@

AUVs operate underwater used to gather spatio-temporal data from a
targeted domain, carrying a scientific payload to survey the water
column or the seafloor. They run at $\sim 1-3$ m/s, can reach depths
from 100--6000m with in-water staying capacity depending on survey
speed, payload sensors and mission design. By deploying AUVs one can
in principle fill gaps and augment data obtained from either ocean
models, earth observing satellite data, or fixed-location sensors on
buoys due to their speed and mobility. Traditional operations and
surveys with AUVs are usually limited to planning samples along fixed
transects or survey lines on a spatial grid, pre-programmed by the
human operator to consist of different waypoints on the grid. But by
using on-board algorithms to continuously evaluate, update, and refine
future sampling locations, one makes information gathering adaptive
\citep{das11b}.  The space of sampling opportunities is still limited
to the spatial grid, but now there is much more flexibility in
automatic adaption.  For instance, an AUV could reconstruct or modify
a survey line based on what temperatures it is measuring by using
embedded methods in decision-theoretic planning \citep{Rajan12b}.

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution. 
Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19} using state-based autonomous
agents. These approaches typically use a reactive-adaptive scheme,
whereby exploration does not rely on a statistical model of the
environment, but rather adapts based on closing the sensing and
actuation loop (Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise selection of the path
(on the graph) which reduces the information criterion the most, has
been used for adaptive surveys \citep{singh2009efficient,Binney2013}
which focus largely on reducing predictive variance or mutual
information (entropy). Variance and entropy reduction are
independent of the actual data realizations under the assumption of GP models, and the use of data-driven (adaptive) criteria was introduced to include more targeted sampling of regions of scientific interest by
\cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, designs that improve the classification of water masses should be rewarded.

@@

The objective is to characterize the river plume, focusing on spatial separation of cold freshwater from the warmer saline waters of the fjord. ESs and EPs are useful starting points to measure the ability to classify water masses, and the goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

>>>>

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem related phenomena such as blooms,
anoxic zones and plumes, all with significant impact both directly and
indirectly on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observability of severe environmental problems such as the recent (May
2018) outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

@@

Our focus in this work is towards spatial characterization of a
frontal system, and more specifically fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) where cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind and temperature differences, this boundary often gets
distorted. Prior knowledge about the location and evolution of these
features is therefore highly uncertain, making deterministic planning
challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $10 m^2$ -- $5 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit scope to cover only lateral spatial
elements. Moreover, for this case, time-effects can be regarded as
static when acquiring AUV data over a short time window and limited
region. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data from a targeted
domain while carrying a range of scientific payloads to survey the
water column.  Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000m with an in-water staying capacity depending on
survey speed, payload sensors and mission design. By deploying AUVs
one can fill gaps in data gathering and augment information obtained
from either ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator.  By using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}.  The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation.  For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution. 
Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} which focus largely on reducing
predictive variance or mutual information (entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumption of GP models, and the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest by \cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, we reward
the designs that improve the classification of water masses, as a
means to characterize the frontal zone.

@@

Our objective is to characterize a river plume, focusing on spatial
separation of cold freshwater from the warmer saline waters of the
fjord. ESs and EPs are useful starting points to measure the ability
to classify water masses, and the goal of experimental sampling is to
improve this characterization, as quantified via changes in the EPs.
